Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2017-12-29 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
Correction

\begin{abstract} 
The aim of this paper is to model flow instabilities \ie{viscous fingering} using a novel method for modeling multi-phase flows in heterogeneous porous with either uniform or random generated permeability distribution. Starting from modelling the Hele-Shaw experimental apparatus, which is mathematically analogous to the two-dimensional flow in porous media problem. The development of finger structures is investigated by taking into account different conditions such as, different domains under uniform or different permeabilities regions, mobility ratios, mesh types. A novel control finite element method (CVFEM) is used to discretize the governing equations that will solve the multi-fluid porous media flow model. The governing multiphase porous media flow equations are solved using an unstructured mesh by introducing adaptivity in later stages. Dynamic mesh adaptivity will not only help to better capture the development of finger but also allow the system to focus the computational effort where is needed, minimizing both the CPU time and the numerical diffusion. It was found that mobility ratio along with the initial and boundary conditions, as well as the use of adaptivity on top of an ustructured mesh allow the user to better describe and visualise the phenomenon.   
\end{abstract}

\begin{keyword} %% keywords here, in the form: keyword \sep keyword
 Multi-fluid flows \sep Porous media \sep Viscous Instabilities \sep Mobility Ratio \sep Preferential flowpath \sep Finite Element Methods \sep Capillary number \sep Unstructured domain \sep Mesh adaptivity.
\end{keyword}

@@

\medskip
In Sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}, finger's formation, growth and branching were triggered by domain heterogeneity represented by permeability differences. Figure~\ref{fig:3DChannel_sat} shows preferential flow pathways at various instances of time. The progress of the saturation front and formation and growth of fingers in time and space can be seen in more details in isosurfaces shown in Fig.~\ref{fig:3DChannel_Isosurf}.

A numerical simulation of the same problem (\ie same geometry, initial mesh, initial and boundary conditions) was also performed using adaptive mesh, Fig.~\ref{fig:3DChannel_sat_adapt}, with resolution ranging from 235242 to 643064 tetrahedral \PNDG[1]{1} element-pairs. Distinct preferable flow pathways are shown  For the same time-steps an inner view on the domain provides with a more detailed description of not only the flow path itself as well as the minor fingers that may be created. At this point resolution plays an important role on capturing these patterns in detail and is also a limititation that needs to be overcome.

In figures \ref{fig:3DChannel_sat_adapt}(a)-(f) an unstructured and adaptive mesh has been introduced in order to achieve higher resolution and focus the computational effort where is needed in he domain. For the same time-steps an inner view on the domain provides with a more detailed description of not only the flow path itself as well as the minor fingers that may be created. At this point resolution plays an important role on capturing these patterns in detail and is also a limititation that needs to be overcome.

\medskip
A quick review to the mechanisms of viscous fingering (\ie shielding, spreading and tip-splitting, \citet{homsy_1987}) is required. In all the $2$D previous cases (sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}), as interfacial surface tension (IFT) becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability. Through \textit{shielding} mechanism, one of these fingers will eventually outgrow the other and, will then spread (\textit{spreading}) to occupy the appropriate width of the domain ahead of it. Following this process, it reaches a width that is again unstable and the pattern repeats. So far shielding and spreading were dominating the flow but, using the current $3$D domain and by applying two types of meshes, despite the resolution restrictions, the conditions and the mechanics of tip-splitting along a front can be qualitatively understood.

%%%%%%%%%%%%%
\section{Conclusions}\label{Section:Conclusion}

\medskip
A number of different cases have been illustrated in Section \ref{section:results}. Starting from testing experimental cases and geometrical domains (Hele-shaw domain) to verify that the numerical model is working and set up some initial benchmarks, to more complicated cases where various scenarios were implied. Observations made on various mobility ratios, geometrical domains and permeability zones as well as on the role of interfacial forces and fingering mechanisms on the developement of the flows.

>>>>

\begin{abstract} 
  The aim of this paper is to model flow instabilities \ie viscous fingering using a novel method for modeling multi-phase flows in heterogeneous porous media. Starting from modelling the Hele-Shaw experimental apparatus, which is mathematically analogous to the two-dimensional flow in porous media problem. The development of finger structures is investigated by taking into account different conditions such as, different domains under uniform or different permeabilities regions, mobility ratios, mesh types. A novel control finite element method (CVFEM) is used to discretize the governing equations that will solve the multi-fluid porous media flow model. The governing multiphase porous media flow equations are solved using an unstructured mesh by introducing adaptivity in later stages. Dynamic mesh adaptivity will not only help to better capture the development of finger but also allow the system to focus the computational effort where is needed, minimizing both the CPU time and the numerical diffusion. It was found that mobility ratio along with the initial and boundary conditions, as well as the use of adaptivity on top of an ustructured mesh allow the user to better describe and visualise the phenomenon.   \red{KOSTAS: Abstracts are divided into 4 parts: (1) Motivation (1-2 sentences); (2) Main aim of the manuscript highlighting the novelty of the work (you must not use the word novelty ... this should be implicit); (3) summary of the main technologies/methods and; (4) summary of the main findings.}

\end{abstract}

\begin{keyword} %% keywords here, in the form: keyword \sep keyword
 Multi-fluid flows \sep Porous media \sep Viscous Instabilities \sep Mobility Ratio \sep Preferential flowpath \sep Finite Element Methods \sep Capillary number \sep Unstructured domain \sep Mesh adaptivity. \red{KOSTAS: choose 3-5 key-words that better represent the work done.}
\end{keyword}

@@

\medskip
In Sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}, finger's formation, growth and branching were triggered by domain heterogeneity represented by permeability differences. Figure~\ref{fig:3DChannel_sat} shows preferential flow pathways at various instances of time. The progress of the saturation front and formation and growth of fingers in time and space can be seen in more details in isosurfaces (Fig.~\ref{fig:3DChannel_Isosurf}.

A numerical simulation of the same problem (\ie same geometry, initial mesh, initial and boundary conditions) was also performed using adaptive mesh, Fig.~\ref{fig:3DChannel_sat_adapt}, with resolution ranging from 235242 to 643064 tetrahedral \PNDG[1]{1} element-pairs (Fig.~\ref{fig:3DChannel_severalfields_adapt}a). Distinct preferable flow pathways can be shown (Fig.~\ref{fig:3DChannel_severalfields_adapt}c), in particular through the central channel (see streamlines in Fig.~\ref{fig:3DChannel_severalfields_adapt}b) and at the top of the domain.

%For the same time-steps an inner view on the domain provides with a more detailed description of not only the flow path itself as well as the minor fingers that may be created. At this point resolution plays an important role on capturing these patterns in detail and is also a limititation that needs to be overcome.
%In figures \ref{fig:3DChannel_sat_adapt}(a)-(f) an unstructured and adaptive mesh has been introduced in order to achieve higher resolution and focus the computational effort where is needed in he domain. For the same time-steps an inner view on the domain provides with a more detailed description of not only the flow path itself as well as the minor fingers that may be created. At this point resolution plays an important role on capturing these patterns in detail and is also a limititation that needs to be overcome.
%\medskip
%A quick review to the mechanisms of viscous fingering (\ie shielding, spreading and tip-splitting, \citet{homsy_1987}) is required. In all the $2$D previous cases (sections \ref{section:results:setup}-\ref{section:results_hete_fix_adapt}), as interfacial surface tension (IFT) becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability. Through \textit{shielding} mechanism, one of these fingers will eventually outgrow the other and, will then spread (\textit{spreading}) to occupy the appropriate width of the domain ahead of it. Following this process, it reaches a width that is again unstable and the pattern repeats. So far shielding and spreading were dominating the flow but, using the current $3$D domain and by applying two types of meshes, despite the resolution restrictions, the conditions and the mechanics of tip-splitting along a front can be qualitatively understood.

%%%%%%%%%%%%%
\section{Conclusions}\label{Section:Conclusion}
\red{KOSTAS: Conclusions of numerical modelling papers are are often very short and divided into 3 parts: (1) Short (1 paragraph) summary of the paper with focus on the main theoretical aspects that were proved by \underline{your work} in the present manuscript; (2) Summary (1-3 paragraphs) of the main results with the appropriate analysis (\ie how you work has proved the theory); and (3) (optional) 1 paragraph of potential future work. Make sure that (1) and (2) are covered in the present Section.}

%A number of different cases have been illustrated in Section \ref{section:results}. Starting from testing experimental cases and geometrical domains (Hele-shaw domain) to verify that the numerical model is working and set up some initial benchmarks, to more complicated cases where various scenarios were implied. Observations made on various mobility ratios, geometrical domains and permeability zones as well as on the role of interfacial forces and fingering mechanisms on the developement of the flows.

### 2017-12-29 Papers/ViscousInstabilities/article_figure1.tex
Correction

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
\begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.6\textwidth]{./Pics1/3D_Channel/3D_channel_sat_30.pdf} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_60.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_90.pdf} } 
      \hbox{\hspace{2.cm} (a) t=0.29s \hspace{5.cm} (b) t=0.56s \hspace{4.cm} (c) t=0.82s}
      \vspace{0.5cm}
      \hbox{\hspace{.5cm} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_110.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_150.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_490.pdf} }
      \hbox{\hspace{2.cm} (d) t=1.00s \hspace{5.cm} (e) t=1.35s\hspace{4.cm} (f) t=3.50s  } }
\caption{3D channel flow (fixed mesh): saturation front evolving in time and space with preferential flow pathways through 3.50 seconds of numerical simulations. Preferential flow pathway can be readily noticed in these frames, mirroring permeability distribution. The domain contains 368513 \PN[1]{1} elements.}
\label{fig:3DChannel_sat}
\end{figure}
\end{landscape}
\clearpage

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
  \begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.55\textwidth, height=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D30c.pdf} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D60c.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D90c.pdf} }
      \hbox{\hspace{2.0cm} (a) t=0.29s \hspace{6.cm} (b) t=0.56s \hspace{4.cm} (c) t=0.82s}
      \vspace{0.5cm}
      \hbox{\hspace{.5cm} \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D110c.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D150c.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D400c.pdf} }
      \hbox{\hspace{3.cm} (d) t=1.00s \hspace{3.cm} (e) t=1.35s\hspace{4.cm} (f) t=3.50s }}
\caption{3D channel flow (fixed mesh): Isosurfaces for wetted phase saturatioh ranging from 0.45 and 0.60 at the same instants of time of Fig.~\ref{fig:3DChannel_sat}. Preferential flow pathway can be readily noticed in (b) and (f). Fingers' formation and growth can be clearly noticed in (b)-(e). The domain contains 368513 \PN[1]{1} elements.}
\label{fig:3DChannel_Isosurf}
\end{figure}
\end{landscape}
\clearpage

@@

%%%%
%%%%  FIGURE
%%%%
\begin{landscape}
  \begin{figure}[ht] 
    \vbox{\vspace{0.cm}
        \hbox{\hspace{5.cm}
            \includegraphics[width=\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_288_Mesh_b} }
      \vspace{.5cm} 
         \hbox{\hspace{0.cm}
             \includegraphics[width=0.75\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_288_StreamLinesHoriz_b} 
      \hspace{.5cm} 
             \includegraphics[width=0.75\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_288_Isosurface_b} 
            }}
\caption{3D Channel: Text to come}
\label{fig:3DChannel_satvel_XYplanes_adapt}
\end{figure}
\end{landscape}
\clearpage

>>>>

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
\begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.6\textwidth]{./Pics1/3D_Channel/3D_channel_sat_30.pdf} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_60.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_90.pdf} } 
      \hbox{\hspace{2.cm} (a) t=0.29s \hspace{5.cm} (b) t=0.56s \hspace{4.cm} (c) t=0.82s}
      \vspace{0.5cm}
      \hbox{\hspace{.5cm} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_110.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_150.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_channel_sat_490.pdf} }
      \hbox{\hspace{2.cm} (d) t=1.00s \hspace{5.cm} (e) t=1.35s\hspace{4.cm} (f) t=3.50s  } }
\caption{3D channel flow (fixed mesh): saturation front evolving in time and space with preferential flow pathways through 3.50 seconds of numerical simulations. Preferential flow pathway can be readily noticed in these frames, mirroring permeability distribution. The domain contains 235242 \PNDG[1]{1} elements.}
\label{fig:3DChannel_sat}
\end{figure}
\end{landscape}
\clearpage

%%%
%%% FIGURE XXXXXX
%%%
\begin{landscape}
  \begin{figure}[ht]
  \vbox{\vspace{-.5cm}
      \hbox{\includegraphics[width=.55\textwidth, height=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D30c.pdf} 
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D60c.pdf}
            \includegraphics[width=.5\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D90c.pdf} }
      \hbox{\hspace{2.0cm} (a) t=0.29s \hspace{6.cm} (b) t=0.56s \hspace{4.cm} (c) t=0.82s}
      \vspace{0.5cm}
      \hbox{\hspace{.5cm} \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D110c.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D150c.pdf}
            \includegraphics[width=.45\textwidth]{./Pics1/3D_Channel/3D_Channel_Saturation1Isosurface_D400c.pdf} }
      \hbox{\hspace{3.cm} (d) t=1.00s \hspace{3.cm} (e) t=1.35s\hspace{4.cm} (f) t=3.50s }}
\caption{3D channel flow (fixed mesh): Isosurfaces for wetted phase saturatioh ranging from 0.45 and 0.60 at the same instants of time of Fig.~\ref{fig:3DChannel_sat}. Preferential flow pathway can be readily noticed in (b) and (f). Fingers' formation and growth can be clearly noticed in (b)-(e). The domain contains 235242 \PNDG[1]{1} elements.}
\label{fig:3DChannel_Isosurf}
\end{figure}
\end{landscape}
\clearpage

@@

%%%%
%%%%  FIGURE
%%%%
\begin{landscape}
  \begin{figure}[ht] 
    \vbox{\vspace{0.cm}
        \hbox{\hspace{5.cm}
          \includegraphics[width=\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_288_Mesh_b} }
        \hbox{\hspace{10cm}(a)}
      \vspace{.5cm} 
         \hbox{\hspace{0.cm}
             \includegraphics[width=0.75\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_288_StreamLinesHoriz_b} 
      \hspace{.5cm} 
             \includegraphics[width=0.75\textwidth]{./Pics1/3D_ChannelAdaptive/Test_SlowNewAdapt_288_Isosurface_b} 
         }
        \hbox{\hspace{6cm}(b) \hspace{8cm}(c)}
    }
\caption{3D Channel (adaptive mesh): (a) Wetting fluid saturation overlapped with mesh; (b) streamlines and; (c) isosurfaces. All plots are shown at 3.78s (Fig.~\ref{fig:3DChannel_satvel_XYplanes_adapt}f). }
\label{fig:3DChannel_severalfields_adapt}
\end{figure}
\end{landscape}
\clearpage
